---
title: "Assignment Base RMD"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###### Read Data From File and load Libraries

```{r ValidateRDAFiles}
set.seed(108)

library(dplyr)
library(nnet)
library(MASS)

load(file = "Training.Dataset.rda")
load(file = "Test.Dataset.rda")
load(file = "Validatioin.Dataset.rda")
dim(Training.Dataset)
dim(Test.Dataset)
dim(Validatioin.Dataset)
```

###### Cleaning

Replaced missing data in transfused variables (plasma, platelets, cell, cryo), days in ICU, and drugs given to 0. 

Created final outcome varibale condition severity and dropped columns related to outcome variable (status, condition and cause of death), days in ICU and entryid. 

Removed remaining rows with NA (990 patients dropped)

```{r Minor Cleaning}


baseR.replace      <- function(x) { replace(x, is.na(x), 0) }

colnames(Training.Dataset)[colSums(is.na(Training.Dataset)) > 0]

Training.Dataset$nplasma <- baseR.replace(Training.Dataset$nplasma)
Training.Dataset$nplatelets <- baseR.replace(Training.Dataset$nplatelets)
Training.Dataset$ncell <- baseR.replace(Training.Dataset$ncell)
Training.Dataset$ndaysicu <- baseR.replace(Training.Dataset$ndaysicu)
Training.Dataset$ncryo <- baseR.replace(Training.Dataset$ncryo)
Training.Dataset$bvii <- baseR.replace(Training.Dataset$bvii)

Training.Dataset$condition_severity <- factor(case_when(
  Training.Dataset$condition  == "dead" ~ "Dead",
  Training.Dataset$condition  == "fully dependent, requiring attention day and night" ~ "Dependent",
  Training.Dataset$condition  == "dependent, but not requiring constant attention" ~ "Dependent",
  Training.Dataset$condition  == "some restriction in lifestyle but independent" ~ "Independent",
  Training.Dataset$condition  == "minor symptoms" ~ "Independent",
  Training.Dataset$condition  == "no symptoms" ~ "Independent",
  TRUE ~ "Unknown"
), levels = c("Dead", "Dependent", "Independent"))

drop <- c("entryid","status","condition", "cause", "ndaysicu")
working.dataset = Training.Dataset[,!(names(Training.Dataset) %in% drop)]
dim(working.dataset)
complete.workingdata <- na.omit(working.dataset)
dim(complete.workingdata)
```

###### Step-wise regression 


```{r StepwiseReg}

test <- multinom(condition_severity ~ ., data = working.dataset)
mlogit_model <- multinom(condition_severity ~. ,data =complete.workingdata, maxit = 1000) 

###   Perform Step-Wise Variable Selection
step <- stepAIC(mlogit_model, direction="both")
step$anova
```




###### Dataset with Key features

Dropping all variables not identified as significant by Stepwise regression and dropping rows that still have NA (954).  

```{r FinalData}

final_col <- c("age", "injurytime", "injurytype", "sbp", "rr", "cc", "gcseye", "gcsmotor", "gcs", 
                         "bheadinj", "bchest", "babdomen", "bpelvis", "bpe", "bdvt", "bstroke", "bbleed", "bmi", 
                         "bgi", "bmaint", "ncell", "nplasma", "nplatelets", "ncryo")
final.working <- working.dataset[final_col]

dim(working.dataset)
dim(final.working)

final_complete <- na.omit(final.working)
dim(final_complete)
```